---
title: "Untitled"
author: "Kehinde Usman"
date: "April 10, 2019"
output: html_document
---
## Loading and preprocessing the data
###Load the data file "activity.csv" by using read.csv into Activity Monitoring Data(amd)
```{r load data}
    amd <- read.csv("activity.csv", head=TRUE, na.strings="NA")
    str(amd)
```

###Preprocess the data by converting the data to Year-Month-Day format and ignoring the NA values in steps
```{r processing data}
amd$date <- as.Date(amd$date)
amd_ignore <- subset(amd, !is.na(amd$steps))
```

## What is mean total number of steps taken per day?
###Calculating the total number of steps taken per day and showing the histogram
```{r steps total}
dailystep <- tapply(amd_ignore$steps, amd_ignore$date, sum, na.rm=TRUE)
dailystep <- dailystep[!is.na(dailystep)]

hist(x=dailystep,
     col="blue",
     breaks = 30,
     xlab="Total Daily Steps",
     ylab="Frequency",
     main="The total number of steps daily (without missinf")
```
###Calculate and report the mean and median of the total number of steps taken per day

```{r mean and median}
mean(dailystep)

median(dailystep)

```

## What is the average daily activity pattern?
###Time series plot of the 5-minutes interval and the average number of steps taken, averaged across all days
```{r plot avergae steps}
avg_daily <- tapply(amd_ignore$steps, amd_ignore$interval, mean, na.rm=TRUE)
dailyPattern <- data.frame(interval=as.integer(names(avg_daily)), avg=avg_daily)

with(dailyPattern,
     plot(interval,
          avg,
          col="red",
          type="l",
          xlab="5-minute Intervals",
          ylab="Average Steps in the Interval Across All Days"))
```
###The 5-minutes interval that has the maximum number of steps on average across all the days in the dataset.
```{r maximum steps}
max_steps <- max(dailyPattern$avg)
dailyPattern[dailyPattern$avg == max_steps, ]

```

## Imputing missing values
###The total number of missing values in the dataset
```{r is NA steps}
na_number <- sum(is.na(amd$steps))
na_number
```
###Devise a strategy for filling in all of the missing values in the dataset.
###Since the missing values is quite large, therefore we impute missing values based on average number of steps in particular 5-minutes interval. 

###A new dataset that is equal to the original dataset but with the missing data filled in
```{r missing data}
amd_missing <- amd
amd_misssteps <- is.na(amd_missing$steps)
avg_daily <- tapply(amd_ignore$steps, amd_ignore$interval, mean, na.rm=TRUE)
amd_missing$steps[amd_misssteps] <- avg_daily[as.character(amd_missing$interval[amd_misssteps])]

```
###Histogram showing the total number of steps taken each day.
```{r total steps with missing}
new_dailystep <- tapply(amd_missing$steps, amd_missing$date, sum, na.rm=TRUE)

hist(x=new_dailystep,
     col="blue",
     breaks = 30,
     xlab="Daily Steps",
     ylab="Frequency",
     main="The total number of steps daily (with missing values)")

```
###The mean and median of the total number of steps taken per day
```{r mean and median new step}
mean(new_dailystep)

median(new_dailystep)

```
###From the first part of the assignment which is total number of steps daily without missing values, the mean values is the same with the imputed data set and the median has a small change. 
###One possible explanation is that when we fill the missing data for the intervals, we use means for intervals, so we have more data close or identical to the means, and median is shifted and becomes identical to the mean.
###The impact of imputing missing data on the estimates of the total daily number of steps is also clear: now we have higher frquency counts in the histogram at the center region (close to the mean).

## Are there differences in activity patterns between weekdays and weekends?
###Create a new factor variable in the dataset with two levels-"weekday" and "weekend" indicating whether a given date is weekday or weekend day.
```{r}
# Create variable with date in correct format
amd_missing$RealDate <- as.Date(amd_missing$date, format = "%Y-%m-%d")
# create a variable with weekdays name
amd_missing$weekday <- weekdays(amd_missing$RealDate)
# create a new variable indicating weekday or weekend
amd_missing$DayType <- ifelse(amd_missing$weekday=='Saturday' | amd_missing$weekday=='Sunday', 'weekend','weekday')
# see first 10 values
head(amd_missing, n=10)
```
###Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}

# create table with steps per time across weekdaydays or weekend days
stepPerTime <- aggregate(steps~interval+DayType,data=amd_missing,FUN=mean,na.action=na.omit)

library(lattice)
xyplot(steps ~ interval | DayType,
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of steps",
       type="l",
       lty=1,
       data=stepPerTime)